---
title: "laser-generic disease models from R"
author: "laser-generic tutorials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 4.5,
    dpi = 100
)
stopifnot(getRversion() >= "4.1.0")
```

This tutorial drives the [`laser-generic`](https://pypi.org/project/laser-generic/)
Python package from R using [`reticulate`](https://rstudio.github.io/reticulate/).
Each section sets up and runs one of the six standard disease models in
`laser.generic`, then plots its compartment trajectories.

Requires **R ≥ 4.1**. The first knit will be slow while `reticulate`
fetches `laser-generic` from PyPI; subsequent knits reuse the cached
environment.

## Setup

```{r r-packages}
library(reticulate)
library(ggplot2)
```

Declare the Python dependencies. `reticulate::py_require()` provisions
an ephemeral, cached environment on demand — no manual `pip install`
needed.

```{r py-deps}
py_require(c("laser-generic", "numpy"))
```

Load the shared helpers (scenario factory, plotting, etc.):

```{r helpers}
source("helpers.R")
setup_python()
```

A compact wrapper that does the boilerplate every model needs.
Each disease section below just supplies its own `components` list
and reads back the trajectories.

```{r runner}
run_model <- function(parameters_list, scenario, components_fn) {
    parameters <- PropertySet(reticulate::dict(parameters_list))
    model <- lg()$Model(scenario, parameters)
    model$components <- components_fn(model)
    model$run()
    model
}
```

A small dial of common knobs we'll reuse:

```{r common-params}
NTICKS    <- 365L * 1L     # one year
POP       <- 100000L       # single-node population
INIT_INF  <- 10L           # initial infections
R0        <- 2.5           # basic reproduction number for SIR/SEIR-style sims
INF_DUR   <- 7             # mean infectious days
EXP_DUR   <- 3             # mean exposed days (SEIR / SEIRS)
WAN_DUR   <- 180           # mean waning days (SIRS / SEIRS)
BETA      <- R0 / INF_DUR  # per-day transmission rate
```

## SI — simple infection, no recovery

The classic logistic-growth model: every susceptible eventually becomes
infected. Has only β (no recovery rate).

```{r si}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["I"] <- INIT_INF

SI <- lg()$SI

model <- run_model(
    parameters_list = list(prng_seed = 1L, nticks = NTICKS, beta = 0.05),
    scenario        = scenario,
    components_fn   = function(m) list(
        SI$Susceptible(m),
        SI$Infectious(m),
        SI$Transmission(m)
    )
)

compartments_df(model, c("S", "I")) |>
    plot_compartments(title = "SI", subtitle = "Susceptible → Infectious, no recovery")
```

## SIS — recover and return to susceptible

Recovery is permanent for the agent's *current* infection but doesn't
confer immunity. Settles into an endemic equilibrium driven by the
balance between transmission and recovery.

```{r sis}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["I"] <- INIT_INF

SIS <- lg()$SIS
infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)

model <- run_model(
    parameters_list = list(prng_seed = 2L, nticks = NTICKS, beta = BETA),
    scenario        = scenario,
    components_fn   = function(m) list(
        SIS$Susceptible(m),
        SIS$Infectious(m, infdurdist),
        SIS$Transmission(m, infdurdist)
    )
)

compartments_df(model, c("S", "I")) |>
    plot_compartments(title = "SIS", subtitle = sprintf("R0 = %g, inf_dur = %g d", R0, INF_DUR))
```

## SIR — permanent immunity after recovery

Once recovered, agents leave the susceptible pool entirely. Produces
the classic epidemic curve that exhausts itself once enough of the
population is immune.

```{r sir}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["I"] <- INIT_INF
scenario["R"] <- 0L

SIR <- lg()$SIR
infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)

model <- run_model(
    parameters_list = list(prng_seed = 3L, nticks = NTICKS, beta = BETA),
    scenario        = scenario,
    components_fn   = function(m) list(
        SIR$Susceptible(m),
        SIR$Infectious(m, infdurdist),
        SIR$Recovered(m),
        SIR$Transmission(m, infdurdist)
    )
)

compartments_df(model, c("S", "I", "R")) |>
    plot_compartments(title = "SIR", subtitle = sprintf("R0 = %g, inf_dur = %g d", R0, INF_DUR))
```

## SIRS — waning immunity

Recovered agents return to susceptible after a (sampled) waning period.
Produces recurring waves rather than a single epidemic.

```{r sirs}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["I"] <- INIT_INF
scenario["R"] <- 0L

SIRS <- lg()$SIRS
infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)
wandurdist <- .lg_env$laser_dist$normal(loc = WAN_DUR, scale = 30)

model <- run_model(
    parameters_list = list(prng_seed = 4L, nticks = NTICKS, beta = BETA),
    scenario        = scenario,
    components_fn   = function(m) list(
        SIRS$Susceptible(m),
        SIRS$Infectious(m, infdurdist, wandurdist),
        SIRS$Recovered(m, wandurdist),
        SIRS$Transmission(m, infdurdist)
    )
)

compartments_df(model, c("S", "I", "R")) |>
    plot_compartments(title = "SIRS", subtitle = sprintf("wan_dur = %g d", WAN_DUR))
```

## SEIR — adds a latent (exposed) period

The exposed compartment holds individuals who are infected but not yet
infectious. Delays the epidemic onset compared to SIR for the same R0.

```{r seir}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["E"] <- 0L
scenario["I"] <- INIT_INF
scenario["R"] <- 0L

SEIR <- lg()$SEIR
expdurdist <- .lg_env$laser_dist$normal(loc = EXP_DUR, scale = 1.0)
infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)

model <- run_model(
    parameters_list = list(prng_seed = 5L, nticks = NTICKS, beta = BETA),
    scenario        = scenario,
    components_fn   = function(m) list(
        SEIR$Susceptible(m),
        SEIR$Exposed(m, expdurdist, infdurdist),
        SEIR$Infectious(m, infdurdist),
        SEIR$Recovered(m),
        SEIR$Transmission(m, expdurdist)
    )
)

compartments_df(model, c("S", "E", "I", "R")) |>
    plot_compartments(title = "SEIR", subtitle = sprintf("exp_dur = %g d, inf_dur = %g d", EXP_DUR, INF_DUR))
```

## SEIRS — SEIR plus waning immunity

Combines a latent period with waning immunity. The closest of these
six to a realistic measles-style model (modulo demography, which
we cover in `02-customization.Rmd`).

```{r seirs}
seed_everything()

scenario <- make_scenario(pop = POP)
scenario["S"] <- POP - INIT_INF
scenario["E"] <- 0L
scenario["I"] <- INIT_INF
scenario["R"] <- 0L

SEIRS <- lg()$SEIRS
expdurdist <- .lg_env$laser_dist$normal(loc = EXP_DUR, scale = 1.0)
infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)
wandurdist <- .lg_env$laser_dist$normal(loc = WAN_DUR, scale = 30)

model <- run_model(
    parameters_list = list(prng_seed = 6L, nticks = NTICKS, beta = BETA),
    scenario        = scenario,
    components_fn   = function(m) list(
        SEIRS$Susceptible(m),
        SEIRS$Exposed(m, expdurdist, infdurdist),
        SEIRS$Infectious(m, infdurdist, wandurdist),
        SEIRS$Recovered(m, wandurdist),
        SEIRS$Transmission(m, expdurdist)
    )
)

compartments_df(model, c("S", "E", "I", "R")) |>
    plot_compartments(title = "SEIRS", subtitle = "Latent + waning immunity")
```

## What's next

- **Demography.** None of the above include births or deaths. See
  `02-customization.Rmd` for examples of adding `BirthsByCBR`,
  `MortalityByCDR`, `MortalityByEstimator`, and `ConstantPopVitalDynamics`
  to an SEIR base model.
- **Spatial structure.** Every example here uses a single node
  (`M = 1, N = 1`). Pass larger `M` / `N` to `make_scenario()` and a
  per-node population vector to drive a grid; the same component list
  works without modification.
- **Distributions other than `constant()`.** `laser.core.distributions`
  also exposes `normal()`, `lognormal()`, `exponential()`, etc. — swap
  them in for any of the `*durdist` arguments above.
