---
title: "Customizing laser-generic models: adding vital dynamics"
author: "laser-generic tutorials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 4.5,
    dpi = 100
)
stopifnot(getRversion() >= "4.1.0")
```

This tutorial builds on `01-models.Rmd`. We start with a baseline SEIR model
and progressively layer in the four vital-dynamics components from
`laser.generic.vitaldynamics`:

| Component | What it adds |
|---|---|
| `ConstantPopVitalDynamics` | Deaths that immediately recycle into newborn agents — population stays exactly constant. |
| `BirthsByCBR` | Per-tick Poisson births driven by a crude birth rate and an age pyramid. Population grows. |
| `MortalityByCDR` | Per-tick mortality driven by a crude death rate. Population shrinks unless paired with births. |
| `MortalityByEstimator` | Age-specific mortality sampled from a survival curve (Kaplan-Meier). Realistic demography. |

Requires **R ≥ 4.1**. See `README.md` for setup.

## Setup

```{r r-packages}
library(reticulate)
library(ggplot2)
```

```{r py-deps}
py_require(c("laser-generic", "numpy"))
```

```{r helpers}
source("helpers.R")
setup_python()
```

A small utility that adds a total-population trajectory on top of
`compartments_df()`:

```{r helpers-extra}
totals_df <- function(model, states) {
    df <- compartments_df(model, states)
    totals <- aggregate(count ~ tick, data = df, sum)
    totals$compartment <- "N (total)"
    rbind(df, totals)
}

# Plot S/R/N and E/I in two stacked panels with **fixed** y-axis ranges so
# the same outbreak heights look identical across every demographic scenario
# below. Without the anchor points, ggplot's facet_wrap(scales = "free_y")
# would auto-rescale each plot to its own data and visual comparison between
# scenarios would be impossible.
plot_seir_with_zoom <- function(df, title = NULL, subtitle = NULL,
                                full_y_max = 120000L,
                                zoom_y_max = 200L) {
    big   <- "S, R, N (full scale)"
    small <- "E, I (zoomed)"
    df$panel <- factor(
        ifelse(df$compartment %in% c("E", "I"), small, big),
        levels = c(big, small)
    )

    # Invisible anchor points: one per panel, at the top of the desired y-range.
    anchors <- data.frame(
        tick  = c(0L, 0L),
        count = c(full_y_max, zoom_y_max),
        panel = factor(c(big, small), levels = c(big, small))
    )

    ggplot2::ggplot() +
        ggplot2::geom_line(
            data = df,
            ggplot2::aes(tick, count, color = compartment),
            linewidth = 0.6
        ) +
        ggplot2::geom_blank(data = anchors, ggplot2::aes(tick, count)) +
        ggplot2::facet_wrap(~ panel, ncol = 1, scales = "free_y") +
        ggplot2::labs(
            x = "Day", y = "Agents", color = NULL,
            title = title, subtitle = subtitle
        ) +
        ggplot2::theme_minimal(base_size = 10) +
        ggplot2::theme(
            legend.position = "top",
            strip.text      = ggplot2::element_text(hjust = 0)
        )
}
```

## Common SEIR base

We'll keep the disease parameters fixed across every example so the
contribution of each demographic component is the only thing changing.
Two design choices up front shape what every section below shows:

1. **Endemic-equilibrium initial conditions.** Instead of starting with
   the whole population susceptible — which produces one large outbreak
   that consumes almost every susceptible and then leaves nothing for
   vital dynamics to interact with — we seed the steady-state
   structure of an endemic disease:
   `S = N / R0` and `R = N · (1 - 1 / R0)`, with `E = I = 0`. The
   disease enters the population only via reintroduction.
2. **Periodic reintroduction throughout.** A small custom component
   imports 5 random new infections every 60 days. With no vital
   dynamics, the susceptible pool drains permanently and each
   importation lands on an increasingly immune population. With vital
   dynamics, every section below tells a different story about how
   the `S` and `R` populations evolve and how that drives the
   subsequent disease dynamics.

With those two choices, the differences between sections come from
*demography alone* — not from how the simulation was seeded.

```{r common-params}
NTICKS    <- 365L * 5L      # five years
POP       <- 100000L
R0        <- 5             # high enough that each importation produces a
                            # readable bump in E and I; small enough that
                            # the herd-immunity threshold (1 - 1/R0 = 80%)
                            # leaves room for vital dynamics to perturb it.
INF_DUR   <- 7
EXP_DUR   <- 3
BETA      <- R0 / INF_DUR

# Endemic-equilibrium initial conditions. See the narrative above for why.
INIT_S    <- as.integer(round(POP / R0))
INIT_E    <- 0L
INIT_I    <- 0L
INIT_R    <- POP - INIT_S - INIT_E - INIT_I

# Periodic importation: 5 cases every 60 days, throughout the run.
IMP_PERIOD <- 60L
IMP_COUNT  <- 5L

# The four vital-dynamics components share these rate inputs.
CBR       <- 30.0   # crude birth rate, per 1000 per year (Nigeria-ish)
CDR       <- 15.0   # crude death rate, per 1000 per year
```

A factory for the scenario + parameter set — every section calls this so
the only thing that changes between examples is the components list.

```{r seir-factory}
make_seir <- function(birthrates_arg = NULL,
                     nticks = NTICKS,
                     pop    = POP,
                     beta   = BETA,
                     init_s = INIT_S,
                     init_e = INIT_E,
                     init_i = INIT_I,
                     init_r = INIT_R) {
    scenario <- make_scenario(pop = pop)
    scenario["S"] <- init_s
    scenario["E"] <- init_e
    scenario["I"] <- init_i
    scenario["R"] <- init_r

    parameters <- PropertySet(reticulate::dict(list(
        prng_seed = 7L, nticks = as.integer(nticks), beta = beta
    )))

    # `birthrates` must be passed to Model() up-front so it can size capacity
    # for incoming births. If no births, pass NULL.
    if (is.null(birthrates_arg)) {
        model <- lg()$Model(scenario, parameters)
    } else {
        model <- lg()$Model(scenario, parameters, birthrates = birthrates_arg)
    }
    model
}

seir_components <- function(model) {
    SEIR <- lg()$SEIR
    expdurdist <- .lg_env$laser_dist$normal(loc = EXP_DUR, scale = 1.0)
    infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)
    list(
        SEIR$Susceptible(model),
        SEIR$Exposed(model, expdurdist, infdurdist),
        SEIR$Infectious(model, infdurdist),
        SEIR$Recovered(model),
        SEIR$Transmission(model, expdurdist)
    )
}
```

The Nigeria age pyramid and survival curve we'll need for the births and
age-specific-mortality examples:

```{r demographic-data}
age_data <- read.csv(
    "../notebooks/data/Nigeria-Distribution-2020.csv",
    header = FALSE
)
age_pyramid_yearly <- as.numeric(age_data[1:89, 1])
pyramid <- .lg_env$laser_demog$AliasedDistribution(np()$asarray(age_pyramid_yearly))

survival_raw <- read.csv(
    "../notebooks/data/Nigeria-Survival-2020.csv",
    header = FALSE
)
survival_cdf <- cumsum(as.numeric(survival_raw[1:89, 2]))
survival_estimator <- .lg_env$laser_demog$KaplanMeierEstimator(
    np()$asarray(survival_cdf)
)
```

## The reintroduction component

Every section below adds a small custom component that injects
`IMP_COUNT` random new infections every `IMP_PERIOD` ticks. Without it,
an SEIR sim that starts at endemic equilibrium would damp out almost
immediately (E and I are zero at t = 0) and the demographic comparisons
would all show flat lines. With it, the trajectory after the first few
weeks reflects the interplay between reintroduction and whatever the
vital dynamics are doing to the susceptible and recovered pools.

The class is defined in Python via `reticulate::py_run_string()` and
exposed back to R as `PeriodicImportation`. This is also the canonical
pattern for writing any custom component — it just needs a
`step(tick)` method.

```{r importation-class}
reticulate::py_run_string("
import numpy as np
import laser.core.distributions as dists
from laser.generic.shared import State

class PeriodicImportation:
    '''Inject `count` random new infections every `period` ticks.

    Imported cases skip the Exposed state — they're assumed to have
    incubated elsewhere and arrive already infectious, matching how
    real importation events are typically modeled.
    '''
    def __init__(self, model, period, count, infdurdist,
                 infdurmin=1, start=0, end=None):
        self.model = model
        self.period = int(period)
        self.count = int(count)
        self.infdurdist = infdurdist
        self.infdurmin = int(infdurmin)
        self.start = int(start)
        self.end = int(end if end is not None else model.params.nticks)

    def step(self, tick):
        if tick < self.start or tick >= self.end:
            return
        if (tick - self.start) % self.period != 0:
            return

        people = self.model.people
        nodes  = self.model.nodes
        susceptible = np.nonzero(people.state == State.SUSCEPTIBLE.value)[0]
        if len(susceptible) == 0:
            return

        n = min(self.count, len(susceptible))
        chosen = np.random.choice(susceptible, size=n, replace=False)
        people.state[chosen] = State.INFECTIOUS.value

        # Imported cases get a freshly sampled infectious duration.
        samples = dists.sample_floats(self.infdurdist, np.zeros(n, np.float32))
        samples = np.maximum(np.round(samples), self.infdurmin).astype(
            people.itimer.dtype
        )
        people.itimer[chosen] = samples

        # Bookkeep node-level counts for the next tick.
        inf_by_node = np.bincount(
            people.nodeid[chosen], minlength=nodes.count
        ).astype(nodes.S.dtype)
        nodes.S[tick + 1] -= inf_by_node
        nodes.I[tick + 1] += inf_by_node
")

PeriodicImportation <- reticulate::py$PeriodicImportation
```

A small wrapper so the sections below can just call `make_importation(model)`:

```{r importation-factory}
make_importation <- function(model,
                             period = IMP_PERIOD,
                             count  = IMP_COUNT,
                             start  = 0L) {
    infdurdist <- .lg_env$laser_dist$normal(loc = INF_DUR, scale = 1.5)
    PeriodicImportation(
        model      = model,
        period     = period,
        count      = count,
        infdurdist = infdurdist,
        start      = start
    )
}
```

## Baseline — no vital dynamics

Closed population, periodic reintroduction. Every importation event
shrinks the susceptible pool and grows the recovered pool, and nothing
refills `S`. Over five years that drift permanently moves the system
away from its starting equilibrium: `S` decays, `R` accumulates, and
the size of subsequent importation-triggered flare-ups decays accordingly
within stochastic variation.
Total `N` is flat by construction.

```{r baseline-seir, fig.height = 6}
seed_everything()

model <- make_seir()
model$components <- c(
    seir_components(model),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = "SEIR baseline (no vital dynamics)",
        subtitle = "Closed population + periodic importation; S drifts toward zero"
    )
```

## Adding `ConstantPopVitalDynamics`

Recycles deceased agents into newborns at the specified rate (per 1000
per year), keeping total population exactly constant. Useful when you
care about birth-cohort flow but want to avoid sizing capacity for
indefinite growth. Because newborns enter as susceptible while the
deceased come predominantly from `R` (which holds most of the
population at equilibrium), this component replenishes `S` from `R` at
the recycle rate — enough to keep importations producing visible
activity throughout the five years.

```{r constant-pop, fig.height = 6}
seed_everything()

recycle_rates <- ValuesMap()$from_scalar(CDR, as.integer(NTICKS), 1L)

model <- make_seir()
model$components <- c(
    seir_components(model),
    list(.lg_env$laser_vd$ConstantPopVitalDynamics(model, recycle_rates, dobs = TRUE)),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = sprintf("SEIR + ConstantPopVitalDynamics (rate = %g / 1000 / yr)", CDR),
        subtitle = "Total N held fixed; recycle-on-death replenishes susceptibles"
    )
```

## Adding `BirthsByCBR`

Adds Poisson-distributed newborns each tick driven by a crude birth
rate. With no offsetting mortality, the population grows. Every newborn
enters the susceptible compartment, so `S` is replenished at
`CBR × N / 1000 / 365` per day. Together with periodic reintroduction
this keeps importation events firing through the full five years and
the susceptible pool trending up rather than down.

`BirthsByCBR` must come last among the *vital-dynamics* components so
the rest of the tick is computed against today's population, not after
the newborns have joined. The importation component then runs after
that — it writes to `tick + 1`, so its order relative to the rest of
the tick doesn't matter.

```{r births-by-cbr, fig.height = 6}
seed_everything()

birthrates <- ValuesMap()$from_scalar(CBR, as.integer(NTICKS), 1L)

model <- make_seir(birthrates_arg = birthrates)
model$components <- c(
    seir_components(model),
    list(.lg_env$laser_vd$BirthsByCBR(model, birthrates, pyramid, track = TRUE)),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = sprintf("SEIR + BirthsByCBR (CBR = %g / 1000 / yr)", CBR),
        subtitle = "Newborns enter the susceptible pool; N grows over time"
    )
```

## Adding `MortalityByCDR`

Per-tick mortality risk applied uniformly across the population. With
no offsetting births, this drains `R` (the largest compartment) faster
than it drains `S`, so the susceptible *fraction* actually *rises*
even though the absolute population is shrinking. Periodic importation
therefore keeps producing flare-ups slightly higher than the baseline, even
though the total population is falling.

First, mortality alone (no births):

```{r mortality-by-cdr-only, fig.height = 6}
seed_everything()

mortality_rates <- ValuesMap()$from_scalar(CDR, as.integer(NTICKS), 1L)

model <- make_seir()
model$components <- c(
    seir_components(model),
    list(.lg_env$laser_vd$MortalityByCDR(model, mortality_rates)),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = sprintf("SEIR + MortalityByCDR (CDR = %g / 1000 / yr, no births)", CDR),
        subtitle = "Population shrinks but the susceptible fraction can hold steady"
    )
```

Now with births balancing deaths. Both inflows and outflows on every
compartment; the net effect is roughly proportional to CBR − CDR
(positive here), and the susceptible inflow from `BirthsByCBR` keeps
importations producing flare-ups indefinitely.

```{r mortality-by-cdr-with-births, fig.height = 6}
seed_everything()

birthrates      <- ValuesMap()$from_scalar(CBR, as.integer(NTICKS), 1L)
mortality_rates <- ValuesMap()$from_scalar(CDR, as.integer(NTICKS), 1L)

model <- make_seir(birthrates_arg = birthrates)
model$components <- c(
    seir_components(model),
    list(
        .lg_env$laser_vd$MortalityByCDR(model, mortality_rates),
        .lg_env$laser_vd$BirthsByCBR(model, birthrates, pyramid, track = TRUE)
    ),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = sprintf("SEIR + MortalityByCDR + BirthsByCBR (CBR = %g, CDR = %g)", CBR, CDR),
        subtitle = "Births offset deaths; N grows because CBR > CDR"
    )
```

## Adding `MortalityByEstimator`

Age-specific mortality sampled from a Kaplan-Meier survival curve.
Unlike `MortalityByCDR` (uniform per-tick risk), this draws each
agent's date of death from the survival distribution conditional on
their date of birth — which means `BirthsByCBR` must be present with
`track = TRUE` so newborns have a recorded `dob`. The mortality bias
toward older (predominantly recovered) agents drains `R` faster than
it drains `S`, which combined with the steady birth inflow into `S`
is the closest of these setups to a real-world demographic baseline.

```{r mortality-by-estimator, fig.height = 6}
seed_everything()

birthrates <- ValuesMap()$from_scalar(CBR, as.integer(NTICKS), 1L)

model <- make_seir(birthrates_arg = birthrates)
model$components <- c(
    seir_components(model),
    list(
        .lg_env$laser_vd$BirthsByCBR(model, birthrates, pyramid, track = TRUE),
        .lg_env$laser_vd$MortalityByEstimator(model, survival_estimator)
    ),
    list(make_importation(model))
)
invisible(model$run())

totals_df(model, c("S", "E", "I", "R")) |>
    plot_seir_with_zoom(
        title    = "SEIR + BirthsByCBR + MortalityByEstimator",
        subtitle = "Realistic demography: age-specific mortality from a Nigeria survival curve"
    )
```

## Scaling up: measles-style biennial dynamics

The earlier sections all ran at `POP = 100,000` and `R0 = 5` so they
fit comfortably in the introductory budget for a tutorial. At those
settings the per-tick effects of each vital dynamic are visible but
subtle. To see the textbook **biennial outbreak pattern** that
emerged for pre-vaccine measles, we need three changes:

1. **Larger population** (500,000 vs 100,000) — a typical Critical
   Community Size for measles-style dynamics is in the 250-500K range.
   Below that, stochastic die-out between outbreaks dominates and the
   recurring-wave signal vanishes into noise.
2. **Higher R0** (12 vs 5) — within the canonical measles range.
   Each wave consumes nearly all available susceptibles, and the
   trough-to-peak amplitude is large enough to read off the plot.
3. **Faster, more aggressive reintroduction** — 5 cases every six weeks
   instead of every two months, and started a few months in to let
   the equilibrium settle before the dynamics start swinging.

We reuse `PeriodicImportation` from above (no need to redefine the
class) but build the model with the measles-scale parameters and run
for **ten years** so the recurring structure has room to develop.

### Four scenarios, one comparison

We run four configurations under the same measles-scale setup and
stack them on one axis so the differences between vital-dynamics
choices read off at a glance.

```{r importation-comparison, fig.height = 5}
seed_everything()

# Measles-scale overrides (local to this chunk; the smaller-scale
# globals from "Common SEIR base" still apply everywhere else).
M_STICKS   <- 365L * 10L  # 10 years — long enough to resolve recurring waves
M_PERIOD   <- 42L         # every six weeks
M_COUNT    <- 5L
M_START    <- 150L        # let the equilibrium settle before importing kicks in

M_POP      <- 500000L
M_R0       <- 12
M_BETA     <- M_R0 / INF_DUR

# Equilibrium init at measles scale: S ≈ N / R0, most of the rest in R,
# plus a small (~0.08%) endemic prevalence so I isn't exactly zero on day 0.
M_INIT_S   <- as.integer(round(M_POP / M_R0))
M_INIT_I   <- as.integer(round(M_POP * 0.0008))
M_INIT_R   <- M_POP - M_INIT_S - M_INIT_I

birthrates     <- ValuesMap()$from_scalar(CBR, as.integer(M_STICKS), 1L)
recycle_rates  <- ValuesMap()$from_scalar(CDR, as.integer(M_STICKS), 1L)

run_with_importation <- function(extra_components_fn, birthrates_arg = NULL) {
    model <- make_seir(
        birthrates_arg = birthrates_arg,
        nticks         = M_STICKS,
        pop            = M_POP,
        beta           = M_BETA,
        init_s         = M_INIT_S,
        init_i         = M_INIT_I,
        init_r         = M_INIT_R
    )
    model$components <- c(
        seir_components(model),
        extra_components_fn(model),
        list(make_importation(model, period = M_PERIOD, count = M_COUNT, start = M_START))
    )
    invisible(model$run())
    compartments_df(model, c("I"))
}

scenarios <- list(
    "baseline (no vital dynamics)" = list(
        extras = function(m) list(),
        births = NULL
    ),
    "ConstantPopVitalDynamics" = list(
        extras = function(m) list(
            .lg_env$laser_vd$ConstantPopVitalDynamics(m, recycle_rates, dobs = TRUE)
        ),
        births = NULL
    ),
    "BirthsByCBR" = list(
        extras = function(m) list(
            .lg_env$laser_vd$BirthsByCBR(m, birthrates, pyramid, track = TRUE)
        ),
        births = birthrates
    ),
    "BirthsByCBR + MortalityByEstimator" = list(
        extras = function(m) list(
            .lg_env$laser_vd$BirthsByCBR(m, birthrates, pyramid, track = TRUE),
            .lg_env$laser_vd$MortalityByEstimator(m, survival_estimator)
        ),
        births = birthrates
    )
)

all_runs <- do.call(rbind, lapply(names(scenarios), function(label) {
    cfg <- scenarios[[label]]
    df  <- run_with_importation(cfg$extras, birthrates_arg = cfg$births)
    df$scenario <- factor(label, levels = names(scenarios))
    df
}))

ggplot(all_runs, aes(tick, count, color = scenario)) +
    geom_vline(xintercept = M_START, linetype = "dashed",
               color = "grey50", linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    labs(
        title    = sprintf(
            "Measles-scale SEIR (POP = %s, R0 = %g): periodic importation every %d days",
            format(M_POP, big.mark = ","), M_R0, M_PERIOD
        ),
        subtitle = "Recurring waves emerge only when vital dynamics replenish susceptibles",
        x = "Day", y = "Infectious", color = NULL
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "top")
```

What the four lines say:

- **baseline (no vital dynamics).** Without new susceptibles entering
  the population, the initial endemic prevalence damps out and every
  subsequent importation lands on an immune population. The flat zero
  line is the absence of vital dynamics.
- **`ConstantPopVitalDynamics`.** Recycle-on-death continuously feeds
  the susceptible compartment, so importations periodically reignite
  the disease. Waves are stochastic and unevenly spaced because the
  recycle rate (CDR = 15 / 1000 / yr) is lower than the births case
  below.
- **`BirthsByCBR`.** Births at CBR = 30 / 1000 / yr build the
  susceptible pool roughly twice as fast as `ConstantPop` does. Waves
  arrive at a regular cadence and peak higher because more
  susceptibles accumulate between waves.
- **`BirthsByCBR + MortalityByEstimator`.** Same susceptible inflow,
  but the immune population now ages out via realistic age-specific
  mortality. Recovered adults die and are replaced by susceptible
  newborns at a balance that gives the **cleanest biennial signal**
  of the four — outbreaks at roughly two-year intervals once the
  system reaches its rhythm. This matches the canonical pre-vaccine
  measles pattern.

The dashed grey line marks `M_START` (day 150) so it's clear which
features of the trajectory are driven by importation versus by the
original seeded prevalence.

## Choosing between the components

A rough guide:

| If you need ... | Use |
|---|---|
| The simplest closed-population approximation | nothing (baseline) |
| To keep total population pinned, with cohort turnover | `ConstantPopVitalDynamics` |
| Growth driven by a published CBR | `BirthsByCBR` |
| Uniform mortality, possibly paired with births | `MortalityByCDR` |
| Age-realistic mortality from a survival curve | `BirthsByCBR(track = TRUE)` + `MortalityByEstimator` |

For multi-decade runs with realistic age structure, the
`BirthsByCBR + MortalityByEstimator` pair is the closest to a fielded
demographic model.

## What's next

- **Multi-node scenarios.** Pass `M > 1` and `N > 1` to
  `make_scenario()` and a vector of populations to drive a grid. The
  vital-dynamics components apply per-node.
- **Time-varying rates.** Each rate input above is a `ValuesMap`. Build
  one with `ValuesMap()$from_array(np_array)` (where `np_array` has shape
  `(nticks, n_nodes)`) to make CBR/CDR vary across the simulation horizon.
- **Real survival data.** Substitute another country's survival curve
  by replacing the `Nigeria-Survival-2020.csv` load.
